---
title: "Newton Method HW"
author: "Doug Hemken"
date: "November 7, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

### Derive Elements of the Jacobian Matrix

- $F_1(x,y) = (1.5 - x + xy)(y-1) + (2.25 - x + xy^2)(y^2 - 1) + (2.625 - x +xy^3)(y^3 - 1)$

    - $\frac{\partial F_1}{\partial x}=(-1+y)+(1.5 - x + xy)+(-1+y^2)+(2.25 - x + xy^2)+(-1+y^3)+(2.625 - x + xy^3)$

        $\frac{\partial F_1}{\partial x}=3.375 - 3x + (x+1)(y+y^2+y^3)$
        
    - $\frac{\partial F_1}{\partial y}=x(y-1)+(1.5 - x + xy)+2xy(y^2-1)+2y(2.25 - x + xy^2)+3xy^2(y^3-1)+3y^2(2.625 - x + xy^3)$
    

- $F_2(x,y) = (1.5 - x + xy) x + (2.25 - x+xy^2)(2xy) + (2.625 - x + xy^3)(3y^2x)$

    - $\frac{\partial F_2}{\partial x} = (y-1)x+(1.5-x+xy) +(y^2-1)(2xy)+2y(2.25-x+xy^2) +(y^3-1)(3y^2x)+(2.625-x+xy^3)(3y^2)$
    
    - $\frac{\partial F_2}{\partial y} = x+ (2xy)(2xy)+2x(2.25-x+xy^2) +(3y^2x)(3y^2x)+(2.625-x+xy^3)(6xy)$
    
### Build Functions
Denote $x$ as `x[1]` and $y$ as `x[2]`.

- The main function
    ```{r}
    Fx <- function(z) {
      x <- z[1]
      y <- z[2]
      F1 <- (1.5 - x + x*y)*(y-1) + (2.25 - x + x*y^2)*(y^2 - 1) + (2.625 - x +x*y^3)*(y^3 - 1)
      F2 <- (1.5 - x + x*y)*x + (2.25 - x+x*y^2)*(2*x*y) + (2.625 - x + x*y^3)*(3*y^2*x)
      return(c(F1,F2))
    }
    Fx(c(0,0))
    ```
    
- The Jacobian
    ```{r}
    J <- function(z) {
      x <- z[1]
      y <- z[2]
      J11 <- 3.375 - 3*x + (x+1)*(y+y^2+y^3)
      J12 <- x*(y-1)+(1.5 - x + x*y)+2*x*y*(y^2-1)+2*y*(2.25 - x + x*y^2)+3*x*y^2*(y^3-1)+3*y^2*(2.625 - x + x*y^3)
      J21 <- (y-1)*x+(1.5-x+x*y) +(y^2-1)*(2*x*y)+2*y*(2.25-x+x*y^2) +(y^3-1)*(3*y^2*x)+(2.625-x+x*y^3)*(3*y^2)
      J22 <- x+ (2*x*y)*(2*x*y)+2*x*(2.25-x+x*y^2) +(3*y^2*x)*(3*y^2*x)+(2.625-x+x*y^3)*(6*x*y)
      return(matrix(c(J11,J12,J21,J22), ncol=2))
    }
    J(c(0,0))
    ```
    
### Newton's Method

```{r}
x0 <- c(0,0)
i <- 1
while (norm(as.matrix(Fx(x0)), "F") > 1e-8 && i <= 25) {
  x <- solve(J(x0),-Fx(x0))+x0
  x0 <- x
  i <- i+1
}
x;i
```

Check
```{r}
Fx(x)
```

Well, that was easy. Suppose we start someplace else.

```{r}
x0 <- c(5,5)
i <- 1
while (norm(as.matrix(Fx(x0)), "F") > 1e-8 && i <= 25) {
  x <- solve(J(x0),-Fx(x0))+x0
  x0 <- x
  i <- i+1
}
x;i
```

```{r}
x0 <- c(-1,-1)
i <- 1
while (norm(as.matrix(Fx(x0)), "F") > 1e-8 && i <= 100) {
  x <- solve(J(x0),-Fx(x0))+x0
  x0 <- x
  i <- i+1
}
x;i
```

```{r}
x0 <- c(1,-1)
i <- 1
while (norm(as.matrix(Fx(x0)), "F") > 1e-8 && i <= 100) {
  x <- solve(J(x0),-Fx(x0))+x0
  x0 <- x
  i <- i+1
}
x;i
```

```{r}
x0 <- c(-1,1)
i <- 1
while (norm(as.matrix(Fx(x0)), "F") > 1e-8 && i <= 100) {
  x <- solve(J(x0),-Fx(x0))+x0
  x0 <- x
  i <- i+1
}
x;i
```