---
title: "Newton Method HW"
author: "Doug Hemken"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

### Derive Elements of the Jacobian Matrix

- $F_1(x,y) = (1.5 - x + xy)(y-1) + (2.25 - x + xy^2)(y^2 - 1) + (2.625 - x +xy^3)(y^3 - 1)$

    - $\frac{\partial F_1}{\partial x}=(-1+y)+(1.5 - x + xy)+(-1+y^2)+(2.25 - x + xy^2)+(-1+y^3)+(2.625 - x + xy^3)$

        $\frac{\partial F_1}{\partial x}=3.375 - 3x + (x+1)(y+y^2+y^3)$
        
    - $\frac{\partial F_1}{\partial y}=x(y-1)+(1.5 - x + xy)+2xy(y^2-1)+2y(2.25 - x + xy^2)+3xy^2(y^3-1)+3y^2(2.625 - x + xy^3)$
    

- $F_2(x,y) = (1.5 - x + xy) x + (2.25 - x+xy^2)(2xy) + (2.625 - x + xy^3)(3y^2x)$

    - $\frac{\partial F_2}{\partial x} = (y-1)x+(1.5-x+xy) +(y^2-1)(2xy)+2y(2.25-x+xy^2) +(y^3-1)(3y^2x)+(2.625-x+xy^3)(3y^2)$
    
    - $\frac{\partial F_2}{\partial y} = x+ (2xy)(2xy)+2x(2.25-x+xy^2) +(3y^2x)(3y^2x)+(2.625-x+xy^3)(6xy)$
    
### Build Functions
Denote $x$ as `x[1]` and $y$ as `x[2]`.

- The main function
    ```{r Fx}
    Fx <- function(z) {
      x <- z[1]
      y <- z[2]
      F1 <- (1.5 - x + x*y)*(y-1) + (2.25 - x + x*y^2)*(y^2 - 1) + (2.625 - x +x*y^3)*(y^3 - 1)
      F2 <- (1.5 - x + x*y)*x + (2.25 - x+x*y^2)*(2*x*y) + (2.625 - x + x*y^3)*(3*y^2*x)
      return(c(F1,F2))
    }
    Fx(c(0,0))
    ```
    
- The Jacobian
    ```{r Jacobian}
    J <- function(z) {
      x <- z[1]
      y <- z[2]
      J11 <- 3.375 - 3*x + (x+1)*(y+y^2+y^3)
      J12 <- x*(y-1)+(1.5 - x + x*y)+2*x*y*(y^2-1)+2*y*(2.25 - x + x*y^2)+3*x*y^2*(y^3-1)+3*y^2*(2.625 - x + x*y^3)
      J21 <- (y-1)*x+(1.5-x+x*y) +(y^2-1)*(2*x*y)+2*y*(2.25-x+x*y^2) +(y^3-1)*(3*y^2*x)+(2.625-x+x*y^3)*(3*y^2)
      J22 <- x+ (2*x*y)*(2*x*y)+2*x*(2.25-x+x*y^2) +(3*y^2*x)*(3*y^2*x)+(2.625-x+x*y^3)*(6*x*y)
      return(matrix(c(J11,J12,J21,J22), ncol=2))
    }
    J(c(0,0))
    ```
    
### Newton's Method

```{r newton_def}
newton <- function (Fx, Jx, x0, eps=1e-8, max.iter=25) {
  # Fx = function
  # Jx = Jacobian of Fx
  # x0 = initial guess
  # eps = stopping threshold
  # max.iter = maximum steps
  i <- 1
  while (norm(as.matrix(Fx(x0)), "F") > eps && i <= max.iter) {
    x <- solve(Jx(x0),-Fx(x0))+x0
    x0 <- x
    i <- i+1
  }
  return(list(solution=x0, iter=i))
}
```
```{r newtonuse1}
x0 <- c(0,0)
root <- newton(Fx, J, x0)
root
```

Check
```{r check}
Fx(root[["solution"]])
```

Well, that was easy. Suppose we start someplace else.

```{r newtonuse2}
x0 <- c(5,5)
newton(Fx, J, x0)
```

This won't get there.
```{r newtonuse3}
x0 <- c(-1,-1)
newton(Fx, J, x0, max.iter=100)
```

```{r backtrack}
backtrack <- function (Fx,J,p,x0,alpha0 = 1.0, rho = 0.5, c = 0.2) {
  # Fx = function
  # J  = Jacobian of Fx0
  # p  = search direction
  # x0 = current guess
  # alpha0 = initial alpha
  # rho= proportion to backtrack
  # c  = relaxation parameter
  Fx0 <- Fx(x0)
#  fx0 <- J
  relaxSlope <- c*t(p)%*%(t(J)%*%Fx0)
    
  alpha <- alpha0
  Fxtry <- Fx(x0 + alpha*p)
  condition <- 0.5*(Fx0%*%Fx0) + alpha*relaxSlope[1]

  while (0.5*(Fxtry %*% Fxtry) > condition) {
    alpha = rho*alpha
    Fxtry = Fx(x0 + alpha*p)
    condition <- 0.5*(Fx0%*%Fx0) + alpha*relaxSlope[1]
  }
return(alpha)
}
```

```{r newtonback}
newtonback <- function (Fx, Jx, x0, eps=1e-8, max.iter=25) {
  # Fx = function
  # Jx = Jacobian of Fx
  # x0 = initial guess
  # eps = stopping threshold
  # max.iter = maximum steps
  i <- 1
  while (norm(as.matrix(Fx(x0)), "F") > eps && i <= max.iter) {
    p <- -solve(Jx(x0), Fx(x0))
    J <- Jx(x0)
    alpha <- backtrack(Fx, J, p, x0)
    x <- x0 + alpha*p
    x0 <- x
    i <- i+1
  }
  return(list(solution=x0, iter=i))
}
```
```{r newtonback1}
x0 <- c(0,0)
newtonback(Fx, J, x0)
```
```{r newtonback2}
x0 <- c(5,5)
newton(Fx, J, x0)
```

```{r newtonback3}
x0 <- c(-1,1)
newton(Fx, J, x0)
```
```{r newtonback4}
x0 <- c(-1,-1)
newton(Fx, J, x0, max.iter=100)
```

### Chord Method
Only compute the Jacobian at some fixed interval, not at every step.

```{r chord_def}
chord <- function (Fx, Jx, x0, v=10, eps=1e-8, max.iter=25) {
  # Fx = function
  # Jx = Jacobian of Fx
  # x0 = initial guess
  # v = number of steps to hold Jacobian
  # eps = stopping threshold
  # max.iter = maximum steps
  i <- 1
  while (norm(as.matrix(Fx(x0)), "F") > eps && i <= max.iter) {
    if (i%%v==1) J <- Jx(x0)
    p <- -solve(J, Fx(x0))
    alpha <- backtrack(Fx, J, p, x0)
    x <- x0 + alpha*p
    x0 <- x
    i <- i+1
  }
  return(list(solution=x0, iter=i))
}
```

This will take a few extra steps, because the Jacobian is usually inexact for the current estimate.

```{r chord_use1}
x0 <- c(0,0)
chord(Fx,J,x0)
```
```{r chord_use2}
x0 <- c(5,5)
chord(Fx,J,x0, max.iter=100)
```
```{r chord_use3}
x0 <- c(-1,1)
chord(Fx,J,x0)
```

### Forward Difference Method
In the Forward Difference Method, we replace the Jacobian with a "forward difference" approximation.

```{r forward}
fwddiff <- function(Fx, x0, eps) {
  # Fx = function to difference/differentiate
  # x0 = point at which to calculate the difference
  # eps = step size of the approximation
  n <- length(x0)
  g <- matrix(rep(NA,n^2), ncol=n)
  eye <- diag(n)
  for (i in 1:n) {
    g[,i] <- (Fx(x0+eps*eye[i,]) - Fx(x0))/eps
  }
  return(g)
  # returns an approximate Jacobian
}
```

```{r checkfwd}
x0 <- c(1,1)
fwddiff(Fx, x0, eps=0.001)
J(x0)
```
```{r newtonfwd}
newtonfwd <- function (Fx, x0, fwd.eps = 0.1, eps=1e-8, max.iter=25) {
  # Fx = function
  # x0 = initial guess
  # eps = stopping threshold
  # max.iter = maximum steps
  i <- 1
  while (norm(as.matrix(Fx(x0)), "F") > eps && i <= max.iter) {
    J <- fwddiff(Fx, x0, eps=fwd.eps)
    p <- -solve(J, Fx(x0))
    alpha <- backtrack(Fx, J, p, x0)
    x <- x0 + alpha*p
    x0 <- x
    i <- i+1
  }
  return(list(solution=x0, iter=i))
}
```
```{r fwd_use1}
x0 <- c(0,0)
newtonfwd(Fx,x0)
```
```{r fwd_use2}
x0 <- c(5,5)
newtonfwd(Fx,x0)
```

```{r fwd_use3}
x0 <- c(-1,1)
newtonfwd(Fx,x0)
```

```{r fwd_use4}
x0 <- c(-1,-1)
newtonfwd(Fx,x0, max.iter=100)
```

### Broyden's Method
Iterative updates to the (iverse of) the Jacobian.